Multiple sequence alignment

### 2024-03-27 episodes/msa.Rmd
Complete msa episode; draft of annotation.

In this part, you will use two different alignemnt trimmers, (`TrimAl`)[http://trimal.cgenomics.org/] and (`ClipKIT`)[https://jlsteenwyk.com/ClipKIT/].

@@

First, look at the list of options available with `trimAl`

>>>>

In this part, you will use two different alignment trimmers, [`TrimAl`](http://trimal.cgenomics.org/) and [`ClipKIT`](https://jlsteenwyk.com/ClipKIT/), on the results of `mafft`'s E-INS-i algorithm.

@@

First, load the module and look at the list of options available with `trimAl`. Then

```bash
module load trimal
trimal -h
```

`trimAl` offers a lot of different options. You are going to explore two different: gap threshold (`-gt`) and the heuristic method `-automated1`, which automatically decides between three automated methods, `-gappyout`, `-strict` and `-strictplus`, based on the properties of the alignment. The gap threshold methods removes columns that contain a fraction of sequences lower than the cut-off.

For comparison purposes, you will be adding an html output.

:::::: challenge

## Challenge 2.1: TrimAl with gap threshold

Use `trimAl` to remove positions in the alignment that have more than 40% gaps.

::: solution

```bash
trimal -in rpoB.einsi.aln -out rpoB.einsi.trimalgt.aln -gt 0.6 -htmlout rpoB.einsi.trimalgt.aln.html
```

::::::::::::

The "gap threshold" is actually expressed as fraction of non-gap residues.

::::::::

## Challenge 2.2: TrimAl with automated trimming

Use `trimAl` with the automated heuristic algorithm.

::: solution

```bash
trimal -in rpoB.einsi.aln -out rpoB.einsi.trimalauto.aln -automated1 -htmlout rpoB.einsi.trimalauto.aln.html
```

::::::::::::

## Challenge 2.3: Compare the results

Use `scp` to get the files (both alignments and html files) to your own laptop and visualize the results, by opening the `html` files with your browser and the alignment files with `seaview` or the viewer you used above.

::: solution

On your own laptop, inside the folder where you want to import the files. Replace `username` with your own username.

```bash
scp <username>@rackham.uppmax.uu.se:/proj/g2020004/nobackup/3MK013/<username>/phylogenetics/rpoB.einsi.trimal* .
```

On some OS it is necessary to escape the `*`. If the output says something about `no matches found`, try:

```bash
scp <username>@rackham.uppmax.uu.se:/proj/g2020004/nobackup/3MK013/<username>/phylogenetics/rpoB.einsi.trimal\* .
```

::::::::::::

::::::::::::::::

### Trimming with ClipKIT

[`ClipKIT`](https://jlsteenwyk.com/ClipKIT/) is one of the more recent tool to trim multiple sequence alignments. In a nutshell, it tries to preseve phylogenetically-informative sites, rather than trimming gappy regions. Although it also has multiple options and modes, you will only use the default mode, `smart-gap`.

:::::: challenge

## Challenge 2.4: Use ClipKIT

To get an idea of the modes and options, look at the help of ClipKit:

```bash
clipkit -h
```

Then run ClipKIT, explicitly using the `smart-gap` mode. Compare how much ClipKIT has trimmed the original alignment compared to trimAl.

::: solution

```bash
clipkit rpoB.einsi.aln -m smart-gap -l -o rpoB.einsi.clipkit.aln
```

```output
---------------------
| Output Statistics |
---------------------
Original length: 2043
Number of sites kept: 1543
Number of sites trimmed: 500
Percentage of alignment trimmed: 24.474%

Execution time: 0.057s
```

::::::::::::

::::::::::::::::

### Comparing all the results

Import the data and inspect the three alignments.

:::::: challenge

## Challenge 2.5: Import data and compare results

Use `scp` as above.

::: solution

```bash
scp <username>@rackham.uppmax.uu.se:/proj/g2020004/nobackup/3MK013/<username>/phylogenetics/rpoB.einsi.clipkit.aln .

```

Then use `seaview` or another viewer to visualize and compare results.

::::::::::::

::::::::::::::::

The three alignments on top of each other look like this. Click [on this link](episodes/fig/seaview_trimal_vs_clipkit.png) to better see the figure.

![](episodes/fig/seaview_trimal_vs_clipkit.png){alt='Alignments shown in seaview'}
It is of course difficult to draw conclusions based on this figure, but can you spot some trends? What alignment is more likely to generate good results?
